Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-02 paper.tex
Edits from Ben and Ana

The ``succinct tree sequence'' is an ARG data structure
that is both general (in terms of the types of ancestry that can
be described) and computationally efficient~\citep{Wong2023-efficient}.
Originally developed to facilitate large-scale coalescent
simulations~\citep{Kelleher2016-wk}, the methods have been
extended and applied to forward-time
simulations~\citep{Kelleher2018-xc,Haller2018-tree},
calculation of population genetics statistics~\citep{Ralph2020-efficiently}
and ARG inference~\citep{Kelleher2019-ba,Wohns2022-th}.
The succinct tree sequence is based on a simple tabular representation,
which defines a set of nodes, edges, sites and mutations. A \emph{node}
represents a particular genome, which may be an observed sample
or an inferred genetic ancestor. The genetic inheritance between
a pair of nodes along a segment of genome is defined by
the \emph{edge} $(\ell, r, p, c)$, which states that
child node $c$ inherited its genome from parent node $p$
from left coordinate $\ell$ to right coordinate $r$. A \emph{site}
defines a position on the genome
and the ancestral state (allele) at that site.
A \emph{mutation} records the site and node IDs where a mutation
occurs and the derived state (allele).
In addition to these basic elements of the data model, the \texttt{tskit} library
supports additional tables and fields,
the ability to associate arbitrary metadata with
table rows, and facilities to record provenance
information~\citep{Tskit2023-tskit}.

Tskit supports arbitrarily complex patterns of mutation at a particular
site, and it is useful
to define some terminology to classify them.
A mutation's ``parent''
is the first mutation encountered (at that site) on the path to
root from the mutation's node in the local tree corresponding to
the site's position. If no other mutation is encountered on the
path to root, the mutation's parent is null.
The ``derived state'' of a mutation is the allelic state inherited by nodes
in the subtree rooted at the mutation's node (in the local tree), assuming
there are no subsequent descendant mutations.
The ``inherited state'' of a mutation is the derived state of the
parent mutation, if it exists, or the site's ancestral state otherwise.
A ``recurrent mutation`` is a mutation with a non-null parent,
and a ``reversion`` is a recurrent mutation that reverses the state
change of its parent. For example, if we have two mutations
$a$ and $b$, such that $a$ is the parent of $b$,
with state transitions (inherited state $\rightarrow$ derived state)
$a: A\rightarrow T$ and $b: T\rightarrow A$, we define $b$ as a reversion.

Given the node, edge, site and mutation tables we can
efficiently construct the local
genealogical trees along the genome (arising from recombination)
and perform a range of calculations efficiently by
reasoning about the differences between these local
trees~\citep{Kelleher2016-wk,Ralph2020-efficiently}. These
algorithms have led to performance increases of several orders
of magnitude over the state-of-the-art in a range of
applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,
Ralph2020-efficiently,Baumdicker2022-ep}.
The succinct tree sequence encoding
is also very concise, allowing, for example, for millions of
complete human genomes to potentially be stored in a few gigabytes of
space~\citep{Kelleher2019-ba}.

>>>>

The ``succinct tree sequence'' is an ARG data structure
that is both general (in terms of the types of ancestry that can
be described) and computationally efficient~\citep{Wong2023-efficient}.
Originally developed to facilitate large-scale coalescent
simulations~\citep{Kelleher2016-wk}, the methods have been
extended and applied to forward-time
simulations~\citep{Kelleher2018-xc,Haller2018-tree},
calculation of population genetics statistics~\citep{Ralph2020-efficiently}
and ARG inference~\citep{Kelleher2019-ba,Wohns2022-th}.
The succinct tree sequence is based on a simple tabular representation,
which defines a set of nodes, edges, sites and mutations. A \emph{node}
represents a particular genome, which may be an observed sample
or an inferred genetic ancestor. The genetic inheritance between
a pair of nodes along a segment of genome is defined by
the \emph{edge} $(\ell, r, p, c)$, which states that
child node $c$ inherited its genome from parent node $p$
from left coordinate $\ell$ to right coordinate $r$. A \emph{site}
defines the genomic position at which some variation occurs,
and a \emph{mutation} defines the site, node and derived state
that generates variation.
% JK: Note sure this clarifies much at the representational level?
% (note that this requires that the sequences can be aligned and phased).

Given these tables, we can then efficiently construct the local
genealogical trees along the genome (arising from recombination)
and also perform a range of calculations efficiently by
reasoning about the differences between these local
trees~\citep{Kelleher2016-wk,Ralph2020-efficiently}. These
algorithms have led to performance increases of several orders
of magnitude over the state-of-the-art in a range of
applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,
Ralph2020-efficiently,Baumdicker2022-ep}.
The succinct tree sequence encoding
is also very concise, allowing, for example, for millions of
complete human genomes to potentially be stored in a few gigabytes of
space~\citep{Kelleher2019-ba}.
